---
title: "Sleep Project Final Report"
author: "Solana Tabor, Ramsay, Zane"
format: pdf
engine: knitr
---

## Table of Contents

1.  Introduction
2.  Methods
3.  Results
4.  Conclusion
5.  Appendix

## 1. Introduction

(*paste intro here*)

## 2. Methods

(*paste methods here*)

#### Step by Step Explanation (add if wanted):

**Set up** (*add if wanted*)

## 3. Results

**Descriptive Results**


Descriptive results: Through our upper quartile study on sleep quality we discovered that the top-quartile sleepers had a much longer sleep duration (*275 minutes higher*) as well as lower caffeine (*20% less*) use and less device time (*12 minutes less*)  compared to sleepers who were not in the top-quartile. Unexpectedly, top sleepers also had higher anxiety ratings (*1.06 points higher*), more social media use(*18 minutes higher*), and less physical activity(*11.50 minutes less*). 

The average overall wake time was reported to be around 7:30 AM (*7.665845*) and the average sleep time was around 11:00 PM (*23.26906*). 

```{r circular-sleep-times, echo=FALSE, message=FALSE, warning=FALSE}
library(readxl)
library(dplyr)
library(ggplot2)
library(tidyr)

source("./excel_date_time_utils.R")

step1 <- suppressWarnings(read_excel(
  "./merged_diaries_actigraphy.xlsx",
  sheet = "best_cleaned_analysis_ready"
))

step1$`Sleep Onset Decimal Hour` <- as.numeric(step1$`Sleep Onset Decimal Hour`)
step1$`Sleep Offset Time_Decimal Hour` <- as.numeric(step1$`Sleep Offset Time_Decimal Hour`)

# Circular encoding
step1 <- step1 %>%
  mutate(
    Onset_sin = sin(2 * pi * `Sleep Onset Decimal Hour` / 24),
    Onset_cos = cos(2 * pi * `Sleep Onset Decimal Hour` / 24),
    
    Offset_sin = sin(2 * pi * `Sleep Offset Time_Decimal Hour` / 24),
    Offset_cos = cos(2 * pi * `Sleep Offset Time_Decimal Hour` / 24),
    
    Onset_circ = atan2(Onset_sin, Onset_cos),
    Onset_circ = ifelse(Onset_circ < 0, Onset_circ + 2*pi, Onset_circ),
    
    Offset_circ = atan2(Offset_sin, Offset_cos),
    Offset_circ = ifelse(Offset_circ < 0, Offset_circ + 2*pi, Offset_circ),
    
    start_hour = Onset_circ * 24 / (2*pi),
    end_hour = Offset_circ * 24 / (2*pi)
  )

# Clock view data
sleep_plot_data <- step1 %>%
  select(Onset_circ, Offset_circ) %>%
  filter(!is.na(Onset_circ), !is.na(Offset_circ)) %>%
  mutate(id = row_number()) %>%
  pivot_longer(
    cols = c(Onset_circ, Offset_circ),
    names_to = "type",
    values_to = "angle"
  )

# Clock plot
ggplot(sleep_plot_data, aes(x = angle, y = 1, color = type)) +
  geom_point(size = 3, alpha = 0.7) +
  coord_polar(theta = "x", start = -pi/2) +
  scale_x_continuous(
    limits = c(0, 2*pi),
    breaks = (0:11) * 2*pi/12,
    labels = c("12AM","2AM","4AM","6AM","8AM","10AM","12PM",
               "2PM","4PM","6PM","8PM","10PM")
  ) +
  scale_color_manual(
    values = c("Onset_circ" = "darkblue",
               "Offset_circ" = "darkgoldenrod2"),
    labels = c("Start Time", "End Time")
  ) +
  theme_minimal() +
  theme(
    axis.title = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks = element_blank()
  ) +
  labs(
    title = "Sleep Start and End Times (Clock View)",
    color = ""
  )
```

**Regression Results**

The regression study was based on the independent variables: onset time, offset time, onset variability, offset variability, sleep duration and the dependent variables: alertness, anxiety, wake difficulty, and sleep quality. Due to the recurring problem of heteroscedasticity violating the assumption of homoscedasticity in the linear regression analyses conducted for each dependent variable we used an HC3 heteroskedasticity-robust standard errors test to interpret coefficients more accurately. The results showed that sleep duration is a significant predictor for alertness (*p= 0.0096, β= 0.0023*), indicating that longer sleep duration is associated with higher alertness. This made worked with our assumption that a longer/better sleep leads to a stronger sense of alertness throughout the day. 

```{r alertness-duration-plot, echo=FALSE, message=FALSE, warning=FALSE}
library(readxl)
library(ggplot2)
library(dplyr)

step1 <- suppressWarnings(read_excel(
  "./merged_diaries_actigraphy.xlsx",
  sheet = "best_cleaned_analysis_ready"
))

plot_alertness <- step1 %>%
  select(`Sleep Duration`, `Alertness Rating`) %>%
  filter(!is.na(`Sleep Duration`),
         !is.na(`Alertness Rating`))

ggplot(plot_alertness,
       aes(x = `Sleep Duration`,
           y = `Alertness Rating`)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = TRUE) +
  theme_minimal() +
  labs(
    title = "Sleep Duration Predicting Alertness",
    x = "Sleep Duration",
    y = "Alertness Rating"
  )
```

Another notable result is offset variability as a significant predictor for wake difficulty (*p=0.006, β= -0.720*), indicating that a greater inconsistency in wake time is associated with decreased difficulty. This was particularly interesting as it worked against our assumption that a more consistent wake time would lead to less difficult wake process aka lower wake difficulty ratings. 

```{r wake-offset-variability-plot, echo=FALSE, message=FALSE, warning=FALSE}
library(readxl)
library(ggplot2)
library(dplyr)

step1 <- suppressWarnings(read_excel(
  "./merged_diaries_actigraphy.xlsx",
  sheet = "best_cleaned_analysis_ready"
))

step1$`Sleep Offset Time_Decimal Hour` <- as.numeric(step1$`Sleep Offset Time_Decimal Hour`)

sleep_var <- step1 %>%
  group_by(`Global Participant ID`) %>%
  summarise(
    offset_var = sd(`Sleep Offset Time_Decimal Hour`, na.rm = TRUE),
    .groups = "drop"
  )

plot_wake_offset <- step1 %>%
  left_join(sleep_var, by = "Global Participant ID") %>%
  select(offset_var, `Wake Difficulty Rating`) %>%
  filter(!is.na(offset_var),
         !is.na(`Wake Difficulty Rating`))

ggplot(plot_wake_offset,
       aes(x = offset_var,
           y = `Wake Difficulty Rating`)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = TRUE) +
  theme_minimal() +
  labs(
    title = "Offset Variability Predicting Wake Difficulty",
    x = "Sleep Offset Variability",
    y = "Wake Difficulty Rating"
  )
```

**Correlation Results**

We also looked into correlation between the dependent variables and found significant correlations between alertness & wake difficulty (*r=0.42*), alertness & sleep quality (*r=0.31*), and anxiety & sleep quality (*r=0.30*).
```{r correlation-heatmap, echo=FALSE, message=FALSE, warning=FALSE}
library(readxl)
library(dplyr)
library(Hmisc)
library(corrplot)

step1 <- suppressWarnings(read_excel(
  "./merged_diaries_actigraphy.xlsx",
  sheet = "best_cleaned_analysis_ready"
))

dep_vars <- step1 %>%
  select(
    `Alertness Rating`,
    `Wake Difficulty Rating`,
    `Anxiety Rating`,
    `Sleep Quality`
  ) %>%
  filter(complete.cases(.))

cor_results <- rcorr(as.matrix(dep_vars), type = "pearson")

cor_matrix <- cor_results$r
p_matrix <- cor_results$P

diag(p_matrix) <- 0

corrplot(
  cor_matrix,
  method = "color",
  type = "upper",
  diag = FALSE,
  addCoef.col = "black",
  p.mat = p_matrix,
  sig.level = 0.05,
  insig = "blank"
)
```
**Mixed Effects Results**

The data included 147 nights from 27 participants, within the data we noticed some trends although not statistically significance still peaked our interest. The trends we found were the association of more sleep than usual to a better reported sleep quality as well as a later sleep onset associated with a worse reported sleep quality. Though these are interesting trends it's important to emphasize that no predictors reached statistical significance in this study. The model assumptions were mostly satisfied showing no major variance issues and although there were some slight left-skew residuals the overall model fit is considered acceptable.

**Takeaways**

Sleep duration is the strongest consistent predictor within our study. However, most other sleep variables showed limited predictive powers. Some relationships exist between key outcomes, but casual predictors are weak overall.

```{r}
#| echo: false
#| label: (add label)

#PASTE DATA FOR PLOT HERE


```

## 4. Conclusions/Future Work

(*paste conclusion here*)

## 5. Appendix {.appendix}


