---
title: "Sleep Project Final Report"
author: "Solana Tabor, Ramsay, Zane"
format: pdf
engine: knitr
---

## Table of Contents

1.  Introduction
2.  Methods
3.  Results
4.  Conclusion
5.  Appendix

## 1. Introduction

(*paste intro here*)

## 2. Methods

(*paste methods here*)

#### Step by Step Explanation (add if wanted):

-   **Set up** (*add if wanted*)

## 3. Results

-   **Descriptive Results**

    -   Through our upper quartile study on sleep quality we discovered that the top-quartile sleepers had a much longer sleep duration as well as lower caffeine use and less device time compared to sleepers who were not in the top-quartile. Unexpectedly, top sleepers also had higher anxiety ratings, more social media use, and less physical activity.

-   **Regression Results**

    -   The regression study was based on the independent variables: onset time, offset time, onset variability, offset variability, sleep duration and the dependent variables: alertness, anxiety, wake difficulty, and sleep quality. Due to the recurring problem of heteroscedasticity violating the assumption of homoscedasticity in the linear regression analyses conducted for each dependent variable we used an HC3 heteroskedasticity-robust standard errors test to interpret coefficients more accurately. The results showed that sleep duration is a significant predictor for alertness (*p= 0.0096, β= 0.0023*), indicating that longer sleep duration is associated with higher alertness. Another notable result is offset variability as a significant predictor for wake difficulty (*p=0.006, β= -0.720*), indicating that a greater inconsistency in wake time is associated with decreased difficulty

-   **Correlation Results**

    -   We also looked into correlation between the dependent variables and found significant correlations between alertness & wake difficulty (r=0.42), alertness & sleep quality (*r=0.31*), and anxiety & sleep quality (*r=0.30*).

-   **Mixed Effects Results**

    -   The data included 147 nights from 27 participants, within the data we noticed some trends although not statistically significance still peaked our interest. The trends we found were the association of more sleep than usual to a better reported sleep quality as well as a later sleep onset associated with a worse reported sleep quality. Though these are interesting trends it's important to emphasize that no predictors reached statistical significance in this study. The model assumptions were mostly satisfied showing no major variance issues and although there were some slight left-skew residuals the overall model fit is considered acceptable.

-   **Takeaways**

    -   Sleep duration is the strongest consistent predictor within our study. However, most other sleep variables showed limited predictive powers. Some relationships exist between key outcomes, but casual predictors are weak overall.

```{r}
#| echo: false
#| label: (add label)

#PASTE DATA FOR PLOT HERE


```

## 4. Conclusions/Future Work

(*paste conclusion here*)

## 5. Appendix {.appendix}

```{r}
#| echo: true
#| label: Appendix

#ALL OF OUR DATA

#REGRESSION ANALYSIS
#Data Input ----
library(readxl)
source("./excel_date_time_utils.R")

step1 <- suppressWarnings(read_excel("./merged_diaries_actigraphy.xlsx", sheet = "best_cleaned_analysis_ready"))
View(step1)
step1$`Sleep Onset Decimal Hour` <- as.numeric(step1$`Sleep Onset Decimal Hour`)
step1$`Sleep Offset Time_Decimal Hour` <- as.numeric(step1$`Sleep Offset Time_Decimal Hour`)

#making means for jules----
library(dplyr)

means <- step1 %>%
  group_by(step1$`Global Participant ID`) %>%
  summarise(across(where(is.numeric), mean, na.rm = TRUE))
View(means)

install.packages("openxlsx")
library(openxlsx)

write.xlsx(means, "mean_per_person.xlsx")



#All together vers----

#Building blocks----


install.packages("dplyr")
install.packages("ggplot2")
install.packages("car")
install.packages("corrplot")
install.packages("lmtest")
library(dplyr)
library(ggplot2)
library(car)
library(corrplot)
library(dplyr)
library(lmtest)

#Variability Dataset

sleep_var <- step1 %>%
  group_by(`Global Participant ID`) %>%
  summarise(
    onset_var = sd(`Sleep Onset Decimal Hour`, na.rm = TRUE),
    offset_var = sd(`Sleep Offset Time_Decimal Hour`, na.rm = TRUE),
    mean_alertness = mean(`Alertness Rating`, na.rm = TRUE),
    mean_wake_diff = mean(`Wake Difficulty Rating`, na.rm = TRUE),
    mean_anxiety = mean(`Anxiety Rating`, na.rm = TRUE),
    mean_sleep_quality = mean(`Sleep Quality`, na.rm = TRUE)
  )

step1 <- merge(step1, sleep_var,
               by = "Global Participant ID")

#Alertness Analysis----

#onset & offset
alertness1 <- lm(
  `Alertness Rating` ~
    `Sleep Offset Time_Decimal Hour` +
    `Sleep Onset Decimal Hour`,
  data = step1
)

summary(alertness1)
bptest(alertness1)

#onset, offset, duration
alertness2 <- lm(
  `Alertness Rating` ~
    `Sleep Duration` +
    `Sleep Offset Time_Decimal Hour` +
    `Sleep Onset Decimal Hour`,
  data = step1
)

summary(alertness2)
bptest(alertness2)


#onset, offset, duration, variability
alertness3 <- lm(
  `Alertness Rating` ~
    `Sleep Duration` +
    `Sleep Offset Time_Decimal Hour` +
    `Sleep Onset Decimal Hour` +
    onset_var +
    offset_var,
  data = step1
)

summary(alertness3)
bptest(alertness3)


#Wake Difficulty Analysis----

#onset & offset
wake_diff1 <- lm(
  `Wake Difficulty Rating` ~
    `Sleep Offset Time_Decimal Hour` +
    `Sleep Onset Decimal Hour`,
  data = step1
)

summary(wake_diff1)
bptest(wake_diff1)

#onset, offset, duration
wake_diff2 <- lm(
  `Wake Difficulty Rating` ~
    `Sleep Duration` +
    `Sleep Offset Time_Decimal Hour` +
    `Sleep Onset Decimal Hour`,
  data = step1
)

summary(wake_diff2)
bptest(wake_diff2)


#onset, offset, duration, variability
wake_diff3 <- lm(
  `Wake Difficulty Rating` ~
    `Sleep Duration` +
    `Sleep Offset Time_Decimal Hour` +
    `Sleep Onset Decimal Hour` +
    onset_var +
    offset_var,
  data = step1
)

summary(wake_diff3)
bptest(wake_diff3)
#Anxiety Analysis----

#onset & offset
anxiety1 <- lm(
  `Anxiety Rating` ~
    `Sleep Offset Time_Decimal Hour` +
    `Sleep Onset Decimal Hour`,
  data = step1
)

summary(anxiety1)
bptest(anxiety1)

#onset, offset, duration
anxiety2 <- lm(
  `Anxiety Rating` ~
    `Sleep Duration` +
    `Sleep Offset Time_Decimal Hour` +
    `Sleep Onset Decimal Hour`,
  data = step1
)

summary(anxiety2)
bptest(anxiety2)


#onset, offset, duration, variability
anxiety3 <- lm(
  `Anxiety Rating` ~
    `Sleep Duration` +
    `Sleep Offset Time_Decimal Hour` +
    `Sleep Onset Decimal Hour` +
    onset_var +
    offset_var,
  data = step1
)

summary(anxiety3)
bptest(anxiety3)
#Sleep Quality Analysis----

#onset & offset
sleep_quality1 <- lm(
  `Sleep Quality` ~
    `Sleep Offset Time_Decimal Hour` +
    `Sleep Onset Decimal Hour`,
  data = step1
)

summary(sleep_quality1)
bptest(sleep_quality1)

#onset, offset, duration
sleep_quality2 <- lm(
  `Sleep Quality` ~
    `Sleep Duration` +
    `Sleep Offset Time_Decimal Hour` +
    `Sleep Onset Decimal Hour`,
  data = step1
)

summary(sleep_quality2)
bptest(sleep_quality2)


#onset, offset, duration, variability
sleep_quality3 <- lm(
  `Sleep Quality` ~
    `Sleep Duration` +
    `Sleep Offset Time_Decimal Hour` +
    `Sleep Onset Decimal Hour` +
    onset_var +
    offset_var,
  data = step1
)

summary(sleep_quality3)
bptest(sleep_quality3)



#plots----

#CHECK MULTICOLLINEARITY

vif(model_alertness)

#REGRESSION DIAGNOSTIC PLOTS

plot(model_alertness)

#CORRELATION MATRIX

vars <- step1 %>%
  select(
    `Sleep Duration`,
    `Sleep Offset Time_Decimal Hour`,
    `Sleep Onset Decimal Hour`,
    `Alertness Rating`,
    `Wake Difficulty Rating`,
    `Anxiety Rating`,
    `Sleep Quality`
  )

corrplot(cor(vars, use = "pairwise.complete.obs"),
         method = "circle")

#SCATTERPLOTS WITH REGRESSION LINES

ggplot(step1,
       aes(x = `Sleep Onset Decimal Hour`,
           y = `Alertness Rating`)) +
  geom_point(alpha = .4) +
  geom_smooth(method = "lm") +
  theme_minimal()

ggplot(step1,
       aes(x = `Sleep Offset Time_Decimal Hour`,
           y = `Sleep Quality`)) +
  geom_point(alpha = .4) +
  geom_smooth(method = "lm") +
  theme_minimal()

#VARIABILITY EFFECTS VISUALIZATION

ggplot(sleep_var,
       aes(x = onset_var,
           y = mean_alertness)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm") +
  theme_minimal()

ggplot(sleep_var,
       aes(x = offset_var,
           y = mean_sleep_quality)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm") +
  theme_minimal()

#Pearson Correlation on DV----
install.packages("Hmisc")
install.packages("corrplot")
install.packages("GGally")
library(Hmisc)
library(corrplot)
library(GGally)

#dependent variables
dep_vars <- step1[, c(
  "Alertness Rating",
  "Wake Difficulty Rating",
  "Anxiety Rating",
  "Sleep Quality"
)]

#Pearson correlation matrix
cor_matrix <- cor(dep_vars,
                  use = "pairwise.complete.obs",
                  method = "pearson")

print(cor_matrix)

#Correlations WITH p-values
cor_test_results <- rcorr(as.matrix(dep_vars),
                          type = "pearson")

print(cor_test_results)

#Correlation plot
corrplot(cor_matrix,
         method = "circle",
         type = "upper",
         tl.col = "black",
         tl.cex = 0.9)

#Scatterplot matrix with regression lines
ggpairs(dep_vars)


#running robust standard errors----
library(sandwich)
library(lmtest)

coeftest(alertness3, vcov = vcovHC(alertness3, type = "HC3"))
coeftest(anxiety3, vcov = vcovHC(anxiety3, type = "HC3"))
coeftest(wake_diff3, vcov = vcovHC(wake_diff3, type = "HC3"))
coeftest(sleep_quality3, vcov = vcovHC(sleep_quality3, type = "HC3"))







# graphs for the final presentation----

install.packages("ggplot2")
install.packages("corrplot")
install.packages("broom")
install.packages("dplyr")

library(ggplot2)
library(corrplot)
library(broom)
library(dplyr)

# GRAPH 1: Sleep Duration → Alertness

ggplot(step1,
       aes(x = `Sleep Duration`,
           y = `Alertness Rating`)) +
  geom_point(alpha = .5) +
  geom_smooth(method = "lm", se = TRUE) +
  theme_minimal() +
  labs(title = "Sleep Duration Predicting Alertness",
       x = "Sleep Duration",
       y = "Alertness Rating")


# GRAPH 2: Offset Variability → Wake Difficulty

ggplot(step1,
       aes(x = offset_var,
           y = `Wake Difficulty Rating`)) +
  geom_point(alpha = .5) +
  geom_smooth(method = "lm", se = TRUE) +
  theme_minimal() +
  labs(title = "Wake-Time Variability Predicting Wake Difficulty",
       x = "Wake-Time Variability",
       y = "Wake Difficulty Rating")


# GRAPH 3: Duration → Anxiety

ggplot(step1,
       aes(x = `Sleep Duration`,
           y = `Anxiety Rating`)) +
  geom_point(alpha = .5) +
  geom_smooth(method = "lm", se = TRUE) +
  theme_minimal() +
  labs(title = "Sleep Duration Predicting Anxiety",
       x = "Sleep Duration",
       y = "Anxiety Rating")


# GRAPH 4: Offset Variability → Anxiety

ggplot(step1,
       aes(x = offset_var,
           y = `Anxiety Rating`)) +
  geom_point(alpha = .5) +
  geom_smooth(method = "lm", se = TRUE) +
  theme_minimal() +
  labs(title = "Wake-Time Variability Predicting Anxiety",
       x = "Wake-Time Variability",
       y = "Anxiety Rating")


# GRAPH 5: Duration → Sleep Quality

ggplot(step1,
       aes(x = `Sleep Duration`,
           y = `Sleep Quality`)) +
  geom_point(alpha = .5) +
  geom_smooth(method = "lm", se = TRUE) +
  theme_minimal() +
  labs(title = "Sleep Duration Predicting Sleep Quality",
       x = "Sleep Duration",
       y = "Sleep Quality")


# GRAPH 6: Offset Time → Sleep Quality

ggplot(step1,
       aes(x = `Sleep Offset Time_Decimal Hour`,
           y = `Sleep Quality`)) +
  geom_point(alpha = .5) +
  geom_smooth(method = "lm", se = TRUE) +
  theme_minimal() +
  labs(title = "Wake Time Predicting Sleep Quality",
       x = "Sleep Offset (Decimal Hour)",
       y = "Sleep Quality")


# GRAPH 7: CORRELATION HEATMAP (DEPENDENT VARIABLES)

library(Hmisc)

dep_vars <- step1[, c(
  "Alertness Rating",
  "Wake Difficulty Rating",
  "Anxiety Rating",
  "Sleep Quality"
)]

cor_results <- rcorr(as.matrix(dep_vars))

corrplot(cor_results$r,
         p.mat = cor_results$P,
         sig.level = 0.05,
         insig = "blank",
         method = "circle",
         type = "upper")

```
